Fractional Types

### 2013-02-07 frac.tex
fixed latex
sec3 more or less done

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background: {{langRev}} }

@@

motivate the need for relations

Even though the categorical semantics of {{langRevTF}} is not the main focus
of this paper, we find it helpful to briefly outline an interpretation of
{{langRevTF}} in the category of finite sets and relations. This category is
an example of a \emph{biproduct dagger compact closed category} which are
categories introduced by Abramsky and
Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878} and by
Selinger~\cite{Selinger:2007:DCC:1229185.1229207} to model quantum
computation.\footnote{The other common example of these categories is that of
  finite dimensional Hilbert spaces.} The two main features of these
biproduct dagger compact closed categories are that they support an adjoint
for each morphism (to model reversibility) and that they support higher-order
relation (built from the categorical dagger which is modeled by our
fractional types). The latter is intuitively clear as any relation between
$A$ and $B$ can be represented as a subset of {{A * B}}, i.e., as a
value.

To interpret {{langRevTF}} in this category, we associate a finite set
{{ [[b]] }} with each type $b$ and a relation {{ [[c]] }} over
{{ [[b1]] }} and {{ [[b2]] }} for each combinator {{ c : b1 <-> b2 }}.
The denotation of types is straighforward: the denotation of {{1/b}} 
is an isomorphic copy of the denotation of {{b}}.

@@

%subcode{proof} include main
%@ ~
%@@ (left v) ~~swap+~~ (right v)
%
%@ ~
%@@ (right v) ~~swap+~~ (left v)
%----
%@ ~
%@@ (left v) ~~assocl+~~ (left (left v))
%
%@ ~
%@@ (right (left v)) ~~assocl+~~ (left (right v))
%
%@ ~
%@@ (right (right v)) ~~assocl+~~ (right v)
%----
%@ ~
%@@ (left (left v)) ~~assocr+~~ (left v)
%
%@ ~
%@@ (left (right v)) ~~assocr+~~ (right (left v))
%
%@ ~
%@@ (right v) ~~assocr+~~ (right (right v))
%----
%@ ~
%@@ ((), v) ~~unite~~ v
%
%@ ~
%@@ v ~~uniti~~ ((), v)
%----
%@ ~
%@@ (v1, v2) ~~swap*~~ (v2, v1)
%----
%@ ~
%@@ (v1, (v2, v3)) ~~assocl*~~ ((v1, v2), v3)
%
%@ ~
%@@ ((v1, v2), v3) ~~assocr*~~ (v1, (v2, v3))
%----
%@ ~
%@@ (left v1, v3) ~~dist~~ (left (v1, v3))
%
%@ ~
%@@ (right v2, v3) ~~dist~~ (right (v2, v3))
%----
%@ ~
%@@ (left (v1, v3)) ~~factor~~ (left v1, v3)
%
%@ ~
%@@ (right (v2, v3)) ~~factor~~ (right v2, v3)
%----
%@ ~
%@@ v ~~id~~ v
%
%@ v' ~~c~~ v
%@@ v ~~(sym c)~~ v'
%----
%@ v1 ~~c1~~ v2
%@ v2 ~~c2~~ v3
%@@ v1 ~~(c1(;)c2)~~ v3
%
%@ v ~~c1~~ v'
%@@ (left v) ~~(c1 (+) c2)~~ (left v')
%
%@ v ~~c2~~ v'
%@@ (right v) ~~(c1 (+) c2)~~ (right v')
%----
%@ v1 ~~c1~~ v1'
%@ v2 ~~c2~~ v2'
%@@ (v1, v2) ~~(c1 (*) c2)~~ (v1', v2')
%----
%@ v in [[b]]
%@@ () ~~eta*_b~~ (1/v,v)
%
%@ ~
%@@ (1/v,v) ~~eps*_b~~ ()

\newpage

@@

The language {{langRevTF}} is still logically reversible.

\begin{proposition}[Logical Reversibility]
\label{chx:prop:logrev-tracep}
For all combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and 
{{v2 : b2}} we have {{v2 in (evalF c v1)}} iff
{{v1 in (evalB c v2)}}.
\end{proposition}

Every relation is clearly reversible by simply swapping the input and
output components of every pair, i.e., if {{r = {[ (a,b), ... ]} }}
then the inverse of {{r}} is simply {{ {[ (b,a), ... ]} }}.

\begin{proposition}[Logical Reversibility]
\label{prop:logrev-tracet}
For all combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and 
{{v2 : b2}} we have if {{v2 in (evalF c v1)}} then 
{{v1 in (evalB c v2)}}.
\end{proposition}

An interesting point to ponder is what happens to \emph{information}
in circuits like {{annihilate}}? We return to this point later.

As hinted to in the first section, the semantics of fractional types introduces
a form of non-determinism in the sense that there could be zero, one, or several
values that satisfy the constraints imposed by fractional values. To model
such non-determinism, we specify the semantics using \emph{relations}. In
other words, the results of {{evalF c v}} or {{evalB c v}} are no longer
guaranteed to be a single value: they could be \emph{sets} of values.

The introduction of relations makes sequential composition of
combinators more involved than simply passing the value produced by
the first to the second. The following example illustrates our
notation for the composition of relations which is used in the formal
semantics.

\begin{example}[Composition of Relations]
Consider the two relations:
\[\begin{array}{rcl}                                                                       
R_1 &=& \{ (1,2),(1,3),(2,2) \} \\                                                       
R_2 &=& \{ (2,1),(2,3),(3,1) \}                                                          
\end{array}\]
These relations can be expressed in a functional notation as follows:
\[\begin{array}{rcl}                                                                       
R_1 (1) &=& \{ 2,3 \} \\                                                                 
R_1 (2) &=& \{ 2 \} \\                                                                   
R_1 (3) &=& \emptyset \\
\\ 
R_2 (1) &=& \emptyset \\                                                                 
R_2 (2) &=& \{ 1,3 \} \\                                                                 
R_2 (3) &=& \{ 1 \}                                                                      
\end{array}\]
Using the functional notation, the composition $(R_2 \circ R_1)$ is defined as
follows:
\[\begin{array}{rcl}                                                                   
(R_2 \circ R_1)~(a) &=& \bigcup_{b \in R_1(a)} R_2(b)                                  
\end{array}\]
We calculate the composition:
\[\begin{array}[t]{rcl}                                                                  
(R_2 \circ R_1)~(1) &=& \bigcup_{b \in R_1(1)} R_2(b) \\                               
         &=& R_2(2) \cup R_2(3) \\                                                     
         &=& \{1,3\} \cup \{ 1 \} \\                                                   
         &=& \{ 1,3 \} \\                                                              
(R_2 \circ R_1)~(2) &=& \{ 1,3 \} \\                                                   
(R_2 \circ R_1)~(3) &=& \emptyset                                                      
\end{array}\]
In other words, 
\[
(R_2 \circ R_1) = \{(1,1), (1,3), (2,1), (2,3)\}
\]
\end{example}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background: {{langRev}} }
\label{sec:pi}

@@

In {{langRev}} presented in Sec.~\ref{sec:pi}, we have been able to maintain
the simple idea that a syntactic isomorphism denotes a \emph{bijection}. In
richer languages, e.g., in a Turing complete language that admits
non-termination, isomorphisms necessarily denote \emph{partial
  bijections}~\cite{rc2011,infeffects}. As we demonstrate in this section,
once we make the extension to higher-order functions, we must interpret the
syntactic isomorphisms as \emph{relations}. The intuitive idea, is that the
fission point combinator {{eta*_b : 1 <-> 1/b * b}} must be able to map the
element {{() : 1}} to any pair of the form {{(1/v,v)}} with
{{v:b}}. Similarly, the combinator {{eps*_b : 1/b * b <-> 1}} must be able to
accept pairs {{(1/v1,v2)}} in which the negative information {{1/v1}} does
not match the positive information {{v2}}. On such pairs, {{eps*_b}} is
undefined, i.e., it does not produce {{():1}}. In other words, some of the
combinators are not partial and some produce more than one result for a
particular input, and hence we move to the setting of general binary
relations to interpret our combinators.

This setting is not only consistent with logical reversibility but is
arguably the \emph{right} framework for considering expressive reversible
languages (including the interesting examples of quantum languages).  Indeed,
even though the categorical semantics of {{langRevTF}} is not the main focus
of this paper, we find it helpful to state that the category of finite sets
and relations in which we are going to interpret our language, is an example
of a \emph{biproduct dagger compact closed category} which are categories
introduced by Abramsky and Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878}
and by Selinger~\cite{Selinger:2007:DCC:1229185.1229207} to model quantum
computation.\footnote{The other common example of these categories is that of
  finite dimensional Hilbert spaces.} Two of the most relevant features of
these categories for our purposes are that they support an adjoint for each
morphism (to model reversibility) and that they support higher-order relation
(built from the categorical dagger which is modeled by our fractional
types). The latter is intuitively clear as any relation between $A$ and $B$
can be represented as a subset of {{A * B}}, i.e., as a value.

To interpret {{langRevTF}} in this category, we associate a finite set
{{ [[b]] }} with each type {{b}} and a relation {{ [[c]] }} over
{{ [[b1]] }} and {{ [[b2]] }} for each combinator {{ c : b1 <-> b2 }}.

@@

%subcode{proof} include main
%@ ~
%@@ (left v) ~~swap+~~ (right v)
%
%@ ~
%@@ (right v) ~~swap+~~ (left v)
%----
%@ ~
%@@ (left v) ~~assocl+~~ (left (left v))
%
%@ ~
%@@ (right (left v)) ~~assocl+~~ (left (right v))
%
%@ ~
%@@ (right (right v)) ~~assocl+~~ (right v)
%----
%@ ~
%@@ (left (left v)) ~~assocr+~~ (left v)
%
%@ ~
%@@ (left (right v)) ~~assocr+~~ (right (left v))
%
%@ ~
%@@ (right v) ~~assocr+~~ (right (right v))
%----
%@ ~
%@@ ((), v) ~~unite~~ v
%
%@ ~
%@@ v ~~uniti~~ ((), v)
%----
%@ ~
%@@ (v1, v2) ~~swap*~~ (v2, v1)
%----
%@ ~
%@@ (v1, (v2, v3)) ~~assocl*~~ ((v1, v2), v3)
%
%@ ~
%@@ ((v1, v2), v3) ~~assocr*~~ (v1, (v2, v3))
%----
%@ ~
%@@ (left v1, v3) ~~dist~~ (left (v1, v3))
%
%@ ~
%@@ (right v2, v3) ~~dist~~ (right (v2, v3))
%----
%@ ~
%@@ (left (v1, v3)) ~~factor~~ (left v1, v3)
%
%@ ~
%@@ (right (v2, v3)) ~~factor~~ (right v2, v3)
%----
%@ ~
%@@ v ~~id~~ v
%
%@ v' ~~c~~ v
%@@ v ~~(sym c)~~ v'
%----
%@ v1 ~~c1~~ v2
%@ v2 ~~c2~~ v3
%@@ v1 ~~(c1(;)c2)~~ v3
%
%@ v ~~c1~~ v'
%@@ (left v) ~~(c1 (+) c2)~~ (left v')
%
%@ v ~~c2~~ v'
%@@ (right v) ~~(c1 (+) c2)~~ (right v')
%----
%@ v1 ~~c1~~ v1'
%@ v2 ~~c2~~ v2'
%@@ (v1, v2) ~~(c1 (*) c2)~~ (v1', v2')
%----
%@ v in [[b]]
%@@ () ~~eta*_b~~ (1/v,v)
%
%@ ~
%@@ (1/v,v) ~~eps*_b~~ ()
\end{definition}

The language {{langRevTF}} is still logically reversible. This is intuitively
clear as the relation denoted by a combinator can be used in the reverse
direction by swapping inputs and outputs.

\begin{proposition}[Logical Reversibility]
\label{chx:prop:logrev-tracep}
For all combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and 
{{v2 : b2}} we have {{v1 ~~c~~ v2}} iff {{v2 ~~(sym c)~~ v1}}.
\end{proposition}

@@
